### 2018-04-02 equal_example.tex
initial commit

\begin{tabular}{C{0.66in}C{0.66in}C{0.66in}C{0.66in}C{0.66in}C{0.66in}C{0.66in}}
\toprule
 experiment &  human droplets &  mouse droplets &  nonempty droplets &  human and mouse droplets &  multiplet freq &  twice cross celltype freq \\
\midrule
          1 &            2005 &            2005 &               4000 &                        10 &           0.005 &                      0.005 \\
          2 &            2050 &            2050 &               4000 &                       100 &           0.049 &                      0.050 \\
          3 &            2500 &            2500 &               4000 &                      1000 &           0.425 &                      0.500 \\
\bottomrule
\end{tabular}

### 2018-04-02 paper.tex
initial commit

%% Submissions for peer-review must enable line-numbering 
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need 
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions

\usepackage[colorlinks=true, citecolor=blue, urlcolor=blue, linkcolor=blue]{hyperref}

\usepackage{array}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\setlength\extrarowheight{1pt}

\usepackage{float}
\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{Supplemental file}
\setcounter{topnumber}{12}
\setcounter{bottomnumber}{12}
\setcounter{totalnumber}{12}

\title{Estimating the frequency of multiplets in single-cell RNA sequencing from cell-mixing experiments}

\author[1]{Jesse D. Bloom}
\affil[1]{Fred Hutch Cancer Research Center, Seattle, WA  98109}
\corrauthor[1]{Jesse D. Bloom}{jbloom@fredhutch.org}

% \keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
In single-cell RNA-sequencing, it is important to know the frequency with which the sequenced transcriptomes actually derive from multiple cells.
A common method to estimate this frequency is to mix two different types of cells (e.g., human and mouse), and then determine how often the transcriptomes contain a mix of transcripts from both cell types.
When the two cell types are mixed in equal proportion, the calculation of the multiplet frequency from the frequency of mixed transcriptomes is straightforward.
However, there are no published descriptions of how to calculate the multiplet frequency in the general case when the cell types are mixed unequally.
Here I derive equations to analytically calculate the multiplet frequency from the numbers of observed pure and mixed transcriptomes when two cell types are mixed in arbitrary proportions, under the assumption that the loading of cells into droplets or wells is Poisson.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}

Many methods for single-cell RNA sequencing involve partitioning cells into barcoded droplets~\citep{klein2015droplet,macosko2015highly,zheng2017massively}, wells~\citep{gierahn2017seq}, or combinations of wells~\citep{cao2017comprehensive}.
The fundamental principle of these methods is that the number of possible barcoded partitions exceeds the number of cells.
Therefore, most partitions contain at most one cell if the number of cells per partition is Poisson distributed.
However, some fraction of the partitions contain multiple cells, and estimating this \emph{multiplet frequency} is an important aspect of experimental quality control.
Crucially, for many methods, the actual number of possible partitions is not precisely known, and all that is observed is the number of different barcoded transcriptomes after sequencing (i.e., the number of non-empty partitions).

The most common method to determine the multiplet frequency is to perform the experiment on a mix two types of cells (e.g., human and mouse).
During the analysis of the sequencing results, each partition can be classified as containing exclusively transcripts from one of the two cell types, or a mix of transcripts from both cell types.
Partitions that contain a mix of transcripts must be multiplets.
If the two cell types are mixed in equal proportions and the average number of cells per partition is low (so that most multiplets are doublets), then the multiplet frequency can be estimated as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
The logic is that all the multiplets are doublets, and only half the doublets will have cells of both types (the others will have two cells of the same type).
This approach has been used to estimate the multiplet rate during the initial prototyping of most single-cell RNA sequencing methods~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive}.

However, in some cases the two cell types may be mixed in unequal proportions.
Unequal mixing could arise simply from error during cell counting, but it can also be a desirable aspect of experimental design.
For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add substantially smaller numbers of mouse cells to estimate the multiplet frequency while still mostly obtaining data for human cells.
But when the cells are mixed unequally, it is no longer valid to estimate the multiplet frequency as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
Surprisingly, I could find no published descriptions of the correct way to calculate the multiplet frequency from unequal mixes of two cell types.
Here I remedy this gap in the literature by deriving the equations to compute the multiplet frequency when the cells are mixed in arbitrary proportions under the assumption that the number of cells per partition is Poisson distributed.

\section*{Results and Discussion}
\label{sec:results}

\subsection*{Derivation of multiplet frequency from observed numbers of pure and mixed-cell droplets}
We consider the case in which cells of two types (e.g., human and mouse) are distributed into individual barcoded droplets, although the same logic applies if the cells are distributed into barcoded wells or combinations of wells.
We assume the sequencing data have been analyzed so that each non-empty droplet can be classified as containing at least one cell of type 1, at least one cell of type 2, or cells of both types.
We will refer to the number of droplets in each of these three groupings as $N_1$, $N_2$, and $N_{1,2}$, respectively.
For instance, the 10X \texttt{cellranger} pipeline (version 2.1.1) returns these numbers as the ``Estimated Number of Cell Partitions.''

The only assumption of our analysis is that the number of cells per droplet is Poisson distributed.
Let $\mu_1$ be the average number of cells of type 1 per droplet, and $\mu_2$ be the average number of cells of type 2 per droplet.
The average number of cells of any type per droplet is then $\mu_1 + \mu_2$.
So the probability that a droplet contains at least one cell of any type is
\begin{eqnarray}
\label{eq:Pr1}
\Pr\left(c \ge 1\right) &=& 1 - \Pr\left(c = 0 \right) \nonumber \\
&=& 1 - e^{-\mu_1 - \mu_2}.
\end{eqnarray}
Likewise, the probability that a droplet contains multiple cells of any type (e.g., a multiplet) is
\begin{eqnarray}
\label{eq:Pr2}
\Pr\left(c \ge 2\right) &=& 1 - \Pr\left(c = 0\right) - \Pr\left(c = 1\right) \nonumber \\
&=& 1 - e^{-\mu_1 - \mu_2} - \left(\mu_1 + \mu_2\right) e^{-\mu_1 + \mu_2}.
\end{eqnarray}
The multiplet frequency $M$ is simply the probability that a droplet with at least one cell actually contains multiple cells, which is
\begin{eqnarray}
\label{eq:M}
M &=& \frac{\Pr\left(c \ge 2\right)}{\Pr\left(c \ge 1\right)} \nonumber \\
&=& 1 - \frac{\left(\mu_1 + \mu_2\right) e^{-\mu_1 + \mu_2}}{1 - e^{-\mu_1 - \mu_2}}.
\end{eqnarray}
However, to evaluate this expression for $M$, we need to know the values of $\mu_1$ and $\mu_2$.

We can write down equations for $\mu_1$ and $\mu_2$ by again using the fact that the number of cells per droplet is Poisson distributed.
Specifically, if $N$ is the total number of droplets (empty and non-empty), then the expected number of droplets that have at least one cell of type 1 is $N \times \Pr\left(c_1 \ge 1\right) = N \left(1 - e^{-\mu_1}\right)$.
The observed number of droplets with at least one cell of type 1 is $N_1$, so setting the observed number equal to the expected number gives us an equation for $\mu_1$,
\begin{equation}
N_1 = N \left(1 - e^{-\mu_1}\right).
\end{equation}
We can easily solve this equation for $\mu_1$ to yield
\begin{equation}
\label{eq:mu1}
\mu_1 = -\ln\left(\frac{N - N_1}{N}\right),
\end{equation}
and likewise for $\mu_2$ we have
\begin{equation}
\label{eq:mu2}
\mu_2 = -\ln\left(\frac{N - N_2}{N}\right).
\end{equation}
Equations~\ref{eq:mu1} and \ref{eq:mu2} give us a way to determine the values ($\mu_1$ and $\mu_2$) needed to calculate the multiplet frequency (Equation~\ref{eq:M}) in terms of the experimental observables $N_1$ and $N_2$.
Unfortunately, these two equations also require knowledge of the total (empty and non-empty) number of droplets $N$, which is not directly observable from the sequencing data.

However, we can take advantage of another relationship to calculate $N$.
The fraction of all (empty and non-empty) droplets that contain cells of both type is $\frac{N_{1,2}}{N}$, and we expect this fraction to simply be the product of the probability that a droplet contains at least one cell of type 1 with the probability that a droplet contains at least one cell of type 2, which in mathematical terms can be stated as $\Pr\left(c_1 \ge 1 \land c_2 \ge 1\right) = \Pr\left(c_1 \ge 1\right) \times \Pr\left(c_2 \ge 1\right)$.
Therefore, we have
\begin{equation}
\frac{N_{1,2}}{N} = \frac{N_1}{N} \times \frac{N_2}{N}.
\end{equation}
This equation can be solved to give
\begin{equation}
\label{eq:N}
N = \frac{N_1 N_2}{N_{1,2}},
\end{equation}
which can be completely evaluated in terms of the experimental observables.
So we can use Equations~\ref{eq:mu1}, \ref{eq:mu2}, and \ref{eq:N} to calculate $\mu_1$ and $\mu_2$ in terms of the experimental observables, and then use those results to calculate the multiplet frequency via Equation~\ref{eq:M}.
Therefore, we now have an analytic solution to the multiplet frequency in terms of the three experimental observables.

\subsection*{Implementation and example calculations}

\url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet.ipynb}

\begin{table}[h]
\centering
\input{equal_example}
\caption{\label{tab:equal}Caption}
\end{table}

\begin{table}[h]
\centering
\input{unequal_example}
\caption{\label{tab:equal}Caption}
\end{table}

\section*{Conclusions}

\section*{Methods}
The LaTex source for this paper, the Jupyter notebook that implements the calculations, and all materials associated with the writing and review of the paper are publicly available in a GitHub repository at \url{https://github.com/jbloomlab/multiplet_freq}.

\bibliography{references}

\clearpage

\begin{suppfile}
\caption{\label{suppfile:jupyter_notebook}
A Jupyter notebook that implements the calculations in Python and R functions, and does the calculations for the examples shown in the tables in this paper.
}
\end{suppfile}

\begin{suppfile}
\caption{\label{suppfile:html_notebook}
This file contains an HTML rendering of the Jupyter notebook in Supplemental file~\ref{suppfile:jupyter_notebook}.
}
\end{suppfile}

\end{document}

### 2018-04-02 paper.tex
completed draft for _PeerJ_ and _bioRxiv_

@@
 \begin{abstract}
-In single-cell RNA-sequencing, it is important to know the frequency with which the sequenced transcriptomes actually derive from multiple cells.
-A common method to estimate this frequency is to mix two different types of cells (e.g., human and mouse), and then determine how often the transcriptomes contain a mix of transcripts from both cell types.
+In single-cell RNA-sequencing, it is important to know the frequency at which the sequenced transcriptomes actually derive from multiple cells.
+A common method to estimate this multiplet frequency is to mix two different types of cells (e.g., human and mouse), and then determine how often the transcriptomes contain transcripts from both cell types.
 When the two cell types are mixed in equal proportion, the calculation of the multiplet frequency from the frequency of mixed transcriptomes is straightforward.
-However, there are no published descriptions of how to calculate the multiplet frequency in the general case when the cell types are mixed unequally.
+But surprisingly, there are no published descriptions of how to calculate the multiplet frequency in the general case when the cell types are mixed unequally.
 Here I derive equations to analytically calculate the multiplet frequency from the numbers of observed pure and mixed transcriptomes when two cell types are mixed in arbitrary proportions, under the assumption that the loading of cells into droplets or wells is Poisson.
@@
 Many methods for single-cell RNA sequencing involve partitioning cells into barcoded droplets~\citep{klein2015droplet,macosko2015highly,zheng2017massively}, wells~\citep{gierahn2017seq}, or combinations of wells~\citep{cao2017comprehensive}.
-The fundamental principle of these methods is that the number of possible barcoded partitions exceeds the number of cells.
-Therefore, most partitions contain at most one cell if the number of cells per partition is Poisson distributed.
-However, some fraction of the partitions contain multiple cells, and estimating this \emph{multiplet frequency} is an important aspect of experimental quality control.
-Crucially, for many methods, the actual number of possible partitions is not precisely known, and all that is observed is the number of different barcoded transcriptomes after sequencing (i.e., the number of non-empty partitions).
+As long as the number of possible partitions exceeds the number of cells, then most partitions will contain at most one cell.
+However, some fraction of the non-empty partitions will contain multiple cells, and estimating this \emph{multiplet frequency} is an important aspect of experimental quality control.
 
-The most common method to determine the multiplet frequency is to perform the experiment on a mix two types of cells (e.g., human and mouse).
-During the analysis of the sequencing results, each partition can be classified as containing exclusively transcripts from one of the two cell types, or a mix of transcripts from both cell types.
-Partitions that contain a mix of transcripts must be multiplets.
-If the two cell types are mixed in equal proportions and the average number of cells per partition is low (so that most multiplets are doublets), then the multiplet frequency can be estimated as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
+The most common method to determine the multiplet frequency is to use a mix two types of cells (e.g., human and mouse).
+During the analysis of the sequencing results, each non-empty partition can be identified as containing transcripts from one or both of the two cell types.
+Partitions that contain a substantial number of transcripts from both cell types must be multiplets.
+If the two cell types are mixed equally and the average number of cells per partition is low (so that most multiplets are doublets), then the multiplet frequency can be estimated as simply twice the fraction of non-empty partitions that contain a mix of cell types.
 The logic is that all the multiplets are doublets, and only half the doublets will have cells of both types (the others will have two cells of the same type).
-This approach has been used to estimate the multiplet rate during the initial prototyping of most single-cell RNA sequencing methods~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive}.
+This approach has been used to estimate the multiplet frequency during the prototyping of most single-cell RNA sequencing methods~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive}.
 
@@
 Unequal mixing could arise simply from error during cell counting, but it can also be a desirable aspect of experimental design.
-For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add substantially smaller numbers of mouse cells to estimate the multiplet frequency while still mostly obtaining data for human cells.
+For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add fewer mouse cells so that most of the resulting data is for the human cells.
 But when the cells are mixed unequally, it is no longer valid to estimate the multiplet frequency as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
-Surprisingly, I could find no published descriptions of the correct way to calculate the multiplet frequency from unequal mixes of two cell types.
+Surprisingly, I could find no published descriptions of how to calculate the multiplet frequency from unequal mixes of two cell types.
 Here I remedy this gap in the literature by deriving the equations to compute the multiplet frequency when the cells are mixed in arbitrary proportions under the assumption that the number of cells per partition is Poisson distributed.
@@
 However, we can take advantage of another relationship to calculate $N$.
-The fraction of all (empty and non-empty) droplets that contain cells of both type is $\frac{N_{1,2}}{N}$, and we expect this fraction to simply be the product of the probability that a droplet contains at least one cell of type 1 with the probability that a droplet contains at least one cell of type 2, which in mathematical terms can be stated as $\Pr\left(c_1 \ge 1 \land c_2 \ge 1\right) = \Pr\left(c_1 \ge 1\right) \times \Pr\left(c_2 \ge 1\right)$.
+The fraction of all (empty and non-empty) droplets that contain cells of both types is $\frac{N_{1,2}}{N}$, and we expect this fraction to simply be the product of the probability that a droplet contains at least one cell of type 1 with the probability that a droplet contains at least one cell of type 2, which in mathematical terms can be stated as $\Pr\left(c_1 \ge 1 \land c_2 \ge 1\right) = \Pr\left(c_1 \ge 1\right) \times \Pr\left(c_2 \ge 1\right)$.
 Therefore, we have
@@
 \subsection*{Implementation and example calculations}
+A simple function to perform the calculations described in the previous subsection is implemented in Python and R in the Jupyter notebook found at
+\url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet.ipynb} (see also Supplemental files~\ref{suppfile:jupyter_notebook} and \ref{suppfile:html_notebook}).
+To illustrate the calculations, we used this function to calculate the multiplet frequency for hypothetical data.
 
-\url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet.ipynb}
+First, we considered hypothetical data in which the two types of cells are mixed in equal proportions.
+Prior papers have approximated the multiplet frequency from such experiments as simply twice the fraction of non-empty droplets that contain cells of both types~\citep{klein2015droplet,macosko2015highly,zheng2017massively,cao2017comprehensive}, which is $\frac{N_{1,2}}{N_1 + N_2 - N_{1,2}}$ in the notation defined in the previous subsection.
+Table~\ref{tab:equal} shows that the exact equation derived in the previous subsection gives very similar results to this approximate method as long as the multiplet frequency is low.
+When the multiplet frequency becomes high, the approximate method starts to overestimate the true multiplet frequency, since it fails to account for the fact that some multiplets will contain more than two cells.
 
-\begin{table}[h]
+\begin{table}[b]
 \centering
 \input{equal_example}
-\caption{\label{tab:equal}Caption}
+\caption{\label{tab:equal}
+Multiplet frequencies for three hypothetical experiments in which human and mouse cells are mixed equally.
+The multiplet frequencies calculated using the exact method described here (column \emph{multiplet freq}) are very similar to those obtained simply by multiplying by two the fraction of non-empty droplets that contain cells of both types (column \emph{twice cross celltype freq}).
+However, the two methods are slightly different at higher multiplet frequencies, since the latter method fails to account for multiplets that have more than two cells.}
 \end{table}
 
-\begin{table}[h]
+Next, we considered hypothetical data in which the two types of cells are mixed in unequal proportions.
+Table~\ref{tab:unequal} shows the multiplet frequencies for several such experiments.
+An interesting aspect of the results is that at high multiplet frequencies and very unequal cell proportions, the multiplet frequency is substantially \emph{lower} than the fraction of droplets containing the rarer cell type that contain a mix of both cell types.
+The reason is that multiplets (particularly higher-order ones) become more and more likely to contain at least one cell of the rarer type relative to droplets that contain only one cell.
+For instance, in the final experiment in Table~\ref{tab:unequal}, two thirds of the droplets containing mouse cells have a mix of both cell types, yet less than half the non-empty droplets are multiplets (the multiplet frequency is 0.459).
+This somewhat non-intuitive results illustrates the importance of using the correct mathematical relationship to calculate the multiplet frequency when cell types are mixed unequally.
+
+\begin{table}[t]
 \centering
 \input{unequal_example}
-\caption{\label{tab:equal}Caption}
+\caption{\label{tab:unequal}
+Multiplet frequencies for five hypothetical experiments in which human and mouse cells are mixed unequally.}
 \end{table}
@@
 \section*{Conclusions}
+I have described how to calculate the multiplet frequency in single-cell RNA sequencing experiments in which two cell types are mixed in arbitrary proportions.
+It is important to note that this calculation requires that the sequencing data have already been analyzed to determine whether each partition contains a non-negligible number of transcripts from each cell type, but many common analysis programs (such as the 10X \texttt{cellranger} pipeline) already perform this analysis.
+It is also important to note that the approach in this paper only calculates the multiplet frequency---it does \emph{not} actually identify the multiplets so that they can be removed from downstream analyses.
+For that purpose, other more sophisticated approaches have been developd~\citep{ilicic2016classification,stoeckius2017cell,kang2018multiplexed}.
+Nonetheless, simply calculating the multiplet frequency from the data returned by standard pipelines such as the 10X \texttt{cellranger} is important for many purposes, and the results here enable that to be done regardless of the proportions at which the cell types are mixed.
 
@@
 The LaTex source for this paper, the Jupyter notebook that implements the calculations, and all materials associated with the writing and review of the paper are publicly available in a GitHub repository at \url{https://github.com/jbloomlab/multiplet_freq}.
+The Jupyter notebook is also available in Supplemental file~\ref{suppfile:jupyter_notebook}, and an HTML rendering of that notebook is in Supplemental file~\ref{suppfile:html_notebook}.
 

### 2018-04-02 paper.tex
tweaks for _PeerJ_ and _bioRxiv_ submission

@@
 
-\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
-% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
+%\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+ \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
 
@@
 \subsection*{Derivation of multiplet frequency from observed numbers of pure and mixed-cell droplets}
-We consider the case in which cells of two types (e.g., human and mouse) are distributed into individual barcoded droplets, although the same logic applies if the cells are distributed into barcoded wells or combinations of wells.
-We assume the sequencing data have been analyzed so that each non-empty droplet can be classified as containing at least one cell of type 1, at least one cell of type 2, or cells of both types.
-We will refer to the number of droplets in each of these three groupings as $N_1$, $N_2$, and $N_{1,2}$, respectively.
+Consider the case in which cells of two types (e.g., human and mouse) are distributed into individual barcoded droplets, although the same logic applies if the cells are distributed into barcoded wells or combinations of wells.
+Assume the sequencing data have been analyzed so that each non-empty droplet can be classified as containing at least one cell of type 1, at least one cell of type 2, or cells of both types.
+I will refer to the number of droplets in each of these three groupings as $N_1$, $N_2$, and $N_{1,2}$, respectively.
 For instance, the 10X \texttt{cellranger} pipeline (version 2.1.1) returns these numbers as the ``Estimated Number of Cell Partitions.''
 
-The only assumption of our analysis is that the number of cells per droplet is Poisson distributed.
+The only assumption of the derivation is that the number of cells per droplet is Poisson distributed.
 Let $\mu_1$ be the average number of cells of type 1 per droplet, and $\mu_2$ be the average number of cells of type 2 per droplet.
@@
 \end{eqnarray}
-However, to evaluate this expression for $M$, we need to know the values of $\mu_1$ and $\mu_2$.
+However, evaluating this expression for $M$ requires the values of $\mu_1$ and $\mu_2$.
 
@@
 \end{equation}
-We can easily solve this equation for $\mu_1$ to yield
+This equation is easily solved for $\mu_1$ to yield
 \begin{equation}
@@
 \end{equation}
-and likewise for $\mu_2$ we have
+and likewise for $\mu_2$,
 \begin{equation}
@@
 However, we can take advantage of another relationship to calculate $N$.
-The fraction of all (empty and non-empty) droplets that contain cells of both types is $\frac{N_{1,2}}{N}$, and we expect this fraction to simply be the product of the probability that a droplet contains at least one cell of type 1 with the probability that a droplet contains at least one cell of type 2, which in mathematical terms can be stated as $\Pr\left(c_1 \ge 1 \land c_2 \ge 1\right) = \Pr\left(c_1 \ge 1\right) \times \Pr\left(c_2 \ge 1\right)$.
-Therefore, we have
+The fraction of all (empty and non-empty) droplets that contain cells of both types is $\frac{N_{1,2}}{N}$, and this fraction is simply the product of the probability that a droplet contains at least one cell of type 1 with the probability that a droplet contains at least one cell of type 2, which in mathematical terms can be stated as $\Pr\left(c_1 \ge 1 \land c_2 \ge 1\right) = \Pr\left(c_1 \ge 1\right) \times \Pr\left(c_2 \ge 1\right)$.
+Therefore,
 \begin{equation}
@@
 which can be completely evaluated in terms of the experimental observables.
-So we can use Equations~\ref{eq:mu1}, \ref{eq:mu2}, and \ref{eq:N} to calculate $\mu_1$ and $\mu_2$ in terms of the experimental observables, and then use those results to calculate the multiplet frequency via Equation~\ref{eq:M}.
-Therefore, we now have an analytic solution to the multiplet frequency in terms of the three experimental observables.
+Equations~\ref{eq:mu1}, \ref{eq:mu2}, and \ref{eq:N} can be used to calculate $\mu_1$ and $\mu_2$ in terms of the experimental observables, and those results used to calculate the multiplet frequency via Equation~\ref{eq:M}.
+This provides an analytic solution for the multiplet frequency in terms of the three experimental observables.
 
@@
 \url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet.ipynb} (see also Supplemental files~\ref{suppfile:jupyter_notebook} and \ref{suppfile:html_notebook}).
-To illustrate the calculations, we used this function to calculate the multiplet frequency for hypothetical data.
+To illustrate the calculations, I used this function to calculate the multiplet frequency for hypothetical data.
 
-First, we considered hypothetical data in which the two types of cells are mixed in equal proportions.
+First, consider hypothetical data in which the two types of cells are mixed in equal proportions.
 Prior papers have approximated the multiplet frequency from such experiments as simply twice the fraction of non-empty droplets that contain cells of both types~\citep{klein2015droplet,macosko2015highly,zheng2017massively,cao2017comprehensive}, which is $\frac{N_{1,2}}{N_1 + N_2 - N_{1,2}}$ in the notation defined in the previous subsection.
@@
 
-Next, we considered hypothetical data in which the two types of cells are mixed in unequal proportions.
+Next, consider hypothetical data in which the two types of cells are mixed in unequal proportions.
 Table~\ref{tab:unequal} shows the multiplet frequencies for several such experiments.
@@
 I have described how to calculate the multiplet frequency in single-cell RNA sequencing experiments in which two cell types are mixed in arbitrary proportions.
-It is important to note that this calculation requires that the sequencing data have already been analyzed to determine whether each partition contains a non-negligible number of transcripts from each cell type, but many common analysis programs (such as the 10X \texttt{cellranger} pipeline) already perform this analysis.
+It is important to note that this calculation requires that the sequencing data have already been analyzed to determine whether each partition contains a non-negligible number of transcripts from each cell type, but many common analysis programs (such as the 10X \texttt{cellranger} pipeline) already do this.
 It is also important to note that the approach in this paper only calculates the multiplet frequency---it does \emph{not} actually identify the multiplets so that they can be removed from downstream analyses.

### 2018-08-02 peerj_rebuttal_letter.md
added reviews

---
geometry: margin=3cm
---

# Rebuttal letter for PeerJ

In the letter below, the reviewer and editor comments are in plain text, and our response is *in italics*.

## Editor's comments
`MINOR REVISIONS`

Both reviewers have a few comments that should be straightforward to address.

### *Response*
*Text*

## Reviewer 1 (Eric Brenner) comments

### Basic reporting
- On line 99, “results” should be singular: “result”. Other than that, the writing was easy to read, and I did not notice any other grammar or spelling mistakes.
- Paragraph starting on line 37. Another strong argument for using unequal proportions of cell types when assessing the accuracy of a scRNA-seq technique is that it is more representative of what you’d find inside of an organism. When working with blood or a tissue containing multiple cell types, it’s unlikely that those cell types will exists in equal proportions. Make sure to emphasize this at some point.
- At some point in the Introduction is important to make it clear that other people have intentionally used unequal proportions of cell types in QC analysis of an scRNA-seq technique. For example, see Figure 2F of Rosenberg et al, 2017 (https://www.biorxiv.org/content/biorxiv/early/2017/02/02/105163.full.pdf). However, your work is novel in that it provides a detailed explanation of how to calculate the multiplet frequency in these scenarios.
- Lines 44-47. Give some brief detail about the Poisson distribution for readers who may not be familiar with it.
- A figure with diagrams showing how the scRNA-seq techniques work may be helpful to some readers.
- Instead of calling R code from within Python (which only works for Python 2 and not Python 3, as far as I know) it would be best to just provide two scripts for readers to choose from that each contain the whole analysis pipeline with one script written purely in Python (no R code) and another purely written in R.

### Experimental design
- See my previous comment regarding the Rosenberg et al paper.
- How closely do you expect the number of cells per droplet to follow a Poisson distribution? From my understanding of microfluidics devices, there would be some upper limit to how many cells could physically fit into a single droplet. This concern may be negligible statistically, but it should still be addressed.

### Validity of the findings
no comment

### Comments for the author
no comment

### *Response*
*Add text*

## Reviewer 2 (Peter Sim) comments

### Basic reporting
The article is well-structured and clearly written with appropriate references and easy-to-follow code (including an html rendering of a Python Jupyter Notebook) for recapitulating the tables in the paper.

### Experimental design
The main purpose of this article was to derive a general equation for evaluating the multiplet frequency in a single-cell RNA-Seq experiment (e.g. from a mixed species experiment). Multiplets represent a major problem in single-cell RNA-Seq (and single cell analysis in general), as they often give profiles that resemble hybrid cell types or states, when they actually originate from a spurious mixture of two or more different cells in a single chamber. The author clearly articulates the problem that the current literature only addresses the case that two species are mixed in equal proportion. He provides an important experimental scenario in which this problem is highly relevant – namely the case that an experimenter includes a small number of cells from a different species as a “spike-in” sample during single-cell RNA-Seq for internal evaluation of the multiplet frequency. He then describes a detailed and straightforward derivation, based on simple Poisson statistics, of an equation that accommodates the more general case that the two species are mixed in arbitrary proportions. The derivation is mathematically sound and provides a straightforward way to evaluate multiplet frequencies from experimentally accessible observables. The methodology used to compute example calculations is easy to replicate, because thoroughly commented code is provided.

### Validity of the findings
The findings presented here are valid and useful. The conclusions are well-supported by a straightforward mathematical derivation – an analytical solution to the proposed problem is provided.

### Comments for the author
I have a few questions and comments regarding the potential for more general application of the analysis described here:
1)  In many single-cell RNA-Seq experiments, unsupervised clustering reveals very discrete cell types that are readily distinguishable, but from the same species. The author focuses on the case that two cells from two different species are mixed in arbitrary proportion, but could these results be extended to the case that an arbitrary number of cell types is mixed in varying proportions? As a concrete example, if I am looking at a blood sample, I might expect to readily separate monocytes, B cells, T cells, etc. Could the framework described here be used to calculate an expectation value of, for example, the doublet frequency of each pair of cell types? This may be beyond the scope of the study described here, but may be worth a comment in the manuscript.
2)  Throughout the paper, the author focuses on what I would call “statistical multiplets”. However, in many single-cell RNA-Seq experiments, this is not the only source of multiplets (and in some cases, not even the dominant source). For example, incomplete cellular dissociation could result in sequencing of multiplets simply because the cells were stuck together in the original tissue and remain stuck together during the profiling experiment. In my own work, I have noticed an enrichment in apparent multiplets coming from cell types that I know are interacting with each other in the tissue I am profiling. This makes the application described in 1) particularly important. Having an expectation value for the “statistical multiplet” frequency could provide a framework for evaluating whether or not the observed multiplet frequency for a pair of cell types is higher than one would expect, thereby implying incomplete dissociation. Again, this may be beyond the scope of the study described here, and I leave this to the author's judgement, but I think it would be worth commenting on the distinction between "statistical multiplets" and multiplets arising from other sources.

## *Response*
*Add text*

### 2018-08-04 peerj_rebuttal_letter.md
added `R` notebook

@@
 
-### *Response*
-*Text*
+*Response:
+Thank you for the careful reviews of the paper, and sorry for the long delay in submitting the revision.
+The delay was due to other intervening non-scientific factors.*
 
@@
 - On line 99, “results” should be singular: “result”. Other than that, the writing was easy to read, and I did not notice any other grammar or spelling mistakes.
+
+*Response:
+Thanks for catching this typo, it is has been corrected.*
+
 - Paragraph starting on line 37.
@@
 Make sure to emphasize this at some point.
+
+*Response:
+This is a good point.
+I have added a sentence explaining that in naturally occurring mixtures of cells of different types (such as blood or tissue), the different cell types will usually be present at unequal proportions.*
+
 - At some point in the Introduction is important to make it clear that other people have intentionally used unequal proportions of cell types in QC analysis of an scRNA-seq technique.
-For example, see Figure 2F of Rosenberg et al, 2017 (https://www.biorxiv.org/content/biorxiv/early/2017/02/02/105163.full.pdf).
+For example, see Figure 2F of [Rosenberg et al, 2017](https://www.biorxiv.org/content/biorxiv/early/2017/02/02/105163.full.pdf).
 However, your work is novel in that it provides a detailed explanation of how to calculate the multiplet frequency in these scenarios.
+
+*Response:
+STILL NEEDED*
+
 - Lines 44-47. Give some brief detail about the Poisson distribution for readers who may not be familiar with it.
+
+*Response:
+I have added text explaining that the Poisson distribution is accurate when cells are loaded into droplets randomly and independently.*
+
 - A figure with diagrams showing how the scRNA-seq techniques work may be helpful to some readers.
+
+*Response:
+Given the wealth of different scRNA-seq techniques that can lead to Poisson loading, I have decided against trying to illustrate them graphically.
+I agree that such diagrams are useful, but they are present in most of the references cited in the Introduction.
+My expectation is that this paper will be of interest primarily to experts who are already familiar with the basics of scRNA-seq and/or have read some of the references that explain these methods.*
+
 - Instead of calling R code from within Python (which only works for Python 2 and not Python 3, as far as I know) it would be best to just provide two scripts for readers to choose from that each contain the whole analysis pipeline with one script written purely in Python (no R code) and another purely written in R.
+
+*Response:
+This is a good idea. 
+We have now provided two separate Juypter notebooks, one that performs the calculations purely in R, and one that performs the calculations purely in Python.*
 
@@
 - See my previous comment regarding the Rosenberg et al paper.
+
 - How closely do you expect the number of cells per droplet to follow a Poisson distribution?
@@
 no comment
-
-### *Response*
-*Add text*
 
@@
 I have a few questions and comments regarding the potential for more general application of the analysis described here:
+
 1)  In many single-cell RNA-Seq experiments, unsupervised clustering reveals very discrete cell types that are readily distinguishable, but from the same species.
@@
 This may be beyond the scope of the study described here, but may be worth a comment in the manuscript.
+
 2)  Throughout the paper, the author focuses on what I would call “statistical multiplets”.
@@
 Again, this may be beyond the scope of the study described here, and I leave this to the author's judgement, but I think it would be worth commenting on the distinction between "statistical multiplets" and multiplets arising from other sources.
-
-## *Response*
-*Add text*

### 2018-08-04 paper.tex
added `R` notebook

@@
 For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add fewer mouse cells so that most of the resulting data is for the human cells.
+In addition, when analyzing naturally occurring mixtures of cells of multiple types, the different cell types will usually be present in unequal proportions.
 But when the cells are mixed unequally, it is no longer valid to estimate the multiplet frequency as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
@@
 Here I remedy this gap in the literature by deriving the equations to compute the multiplet frequency when the cells are mixed in arbitrary proportions under the assumption that the number of cells per partition is Poisson distributed.
+This Poisson assumption is accurate when cells are loaded randomly and independently into partitions.
 
-\section*{Results and Discussion}
+\section*{Methods}
+The LaTex source for this paper, the Jupyter notebooks that implement the calculations, and all materials associated with the writing and review of the paper are publicly available in a GitHub repository at \url{https://github.com/jbloomlab/multiplet_freq}.
+The Jupyter notebooks are also available in Supplemental file~\ref{suppfile:jupyter_notebook} and ~\ref{suppfile:jupyter_notebook_R}, and HTML renderings of the notebooks are in Supplemental file~\ref{suppfile:html_notebook} and \ref{suppfile:html_notebook_R}.
+
+\section*{Results}
 \label{sec:results}
@@
 \subsection*{Implementation and example calculations}
-A simple function to perform the calculations described in the previous subsection is implemented in Python and R in the Jupyter notebook found at
-\url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet.ipynb} (see also Supplemental files~\ref{suppfile:jupyter_notebook} and \ref{suppfile:html_notebook}).
+A simple function to perform the calculations described in the previous subsection is implemented in Python in the Jupyter notebook found at
+\url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet.ipynb}, and in R in the Jupyter notebook found at \url{https://github.com/jbloomlab/multiplet_freq/blob/master/calcmultiplet_R.ipynb} (see also Supplemental files~\ref{suppfile:jupyter_notebook}, \ref{suppfile:html_notebook}, ~\ref{suppfile:jupyter_notebook_R}, and \ref{suppfile:html_notebook_R}).
 To illustrate the calculations, I used this function to calculate the multiplet frequency for hypothetical data.
@@
 For instance, in the final experiment in Table~\ref{tab:unequal}, two thirds of the droplets containing mouse cells have a mix of both cell types, yet less than half the non-empty droplets are multiplets (the multiplet frequency is 0.459).
-This somewhat non-intuitive results illustrates the importance of using the correct mathematical relationship to calculate the multiplet frequency when cell types are mixed unequally.
+This somewhat non-intuitive result illustrates the importance of using the correct mathematical relationship to calculate the multiplet frequency when cell types are mixed unequally.
 
@@
 
-\section*{Methods}
-The LaTex source for this paper, the Jupyter notebook that implements the calculations, and all materials associated with the writing and review of the paper are publicly available in a GitHub repository at \url{https://github.com/jbloomlab/multiplet_freq}.
-The Jupyter notebook is also available in Supplemental file~\ref{suppfile:jupyter_notebook}, and an HTML rendering of that notebook is in Supplemental file~\ref{suppfile:html_notebook}.
-
 \bibliography{references}
@@
 \caption{\label{suppfile:jupyter_notebook}
-A Jupyter notebook that implements the calculations in Python and R functions, and does the calculations for the examples shown in the tables in this paper.
+A Jupyter notebook that implements the calculations in Python, and does the calculations for the examples shown in the tables in this paper.
 }
@@
 
+\begin{suppfile}
+\caption{\label{suppfile:jupyter_notebook_R}
+A Jupyter notebook that implements the calculations in R, and does the calculations for the examples shown in the tables in this paper.
+}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{\label{suppfile:html_notebook_R}
+This file contains an HTML rendering of the Jupyter notebook in Supplemental file~\ref{suppfile:jupyter_notebook_R}.
+}
+\end{suppfile}
+
 \end{document}

### 2018-08-04 peerj_rebuttal_letter.md
completed revision

@@
 *Response:
-STILL NEEDED*
+This is a good point.
+I have added a reference to the Rosenberg et al study in the Introduciton.*
 
@@
 Given the wealth of different scRNA-seq techniques that can lead to Poisson loading, I have decided against trying to illustrate them graphically.
-I agree that such diagrams are useful, but they are present in most of the references cited in the Introduction.
+I agree that such diagrams are useful, but they are present in most of the references cited in the first paragraph of the Introduction.
 My expectation is that this paper will be of interest primarily to experts who are already familiar with the basics of scRNA-seq and/or have read some of the references that explain these methods.*
@@
 This is a good idea. 
-We have now provided two separate Juypter notebooks, one that performs the calculations purely in R, and one that performs the calculations purely in Python.*
+I now provide two separate Juypter notebooks, one that performs the calculations purely in R, and one that performs the calculations purely in Python.*
 
@@
 
+*Response:
+This point has been addressed; see above.*
+
 - How closely do you expect the number of cells per droplet to follow a Poisson distribution?
@@
 This concern may be negligible statistically, but it should still be addressed.
+
+*Response:
+This is an excellent point. 
+I am not able to directly quantify how "Poisson" different methods actually are from available data, so I have added several sentences to the Discussion that discuss how factors such as cell clumping could cause deviations from Poisson behavior, and how this might affect the calculations.*
 
@@
 The article is well-structured and clearly written with appropriate references and easy-to-follow code (including an html rendering of a Python Jupyter Notebook) for recapitulating the tables in the paper.
+
+*Response: 
+Thanks for the nice summary.*
 
@@
 
+*Response: 
+Thanks for the nice summary.*
+
 ### Comments for the author
@@
 
+*Response:
+This is an interesting idea.
+My concern would be that in the case of cell-type clustering, the multiplets might affect the clustering and subsequently the cell-type counting.
+In the case of mixing human and mouse cells, the classification is unambiguous in terms of single-species or mixed-species. 
+But in cell type classification, if there are a substantial number of multiplets, they might skew the clustering, thereby changing the numbers of cells of each type that go into the calculation.
+I am not sure how big a problem this would be in practice, but because of this concern I am reluctant to suggest the approach be applied in such a scenario without extensive further investigation.
+For this reason, I have simply included sentences to the Discussion that emphasize that the method is restricted to cases where the cell types have already been classified.*
+
 2)  Throughout the paper, the author focuses on what I would call “statistical multiplets”.
@@
 Again, this may be beyond the scope of the study described here, and I leave this to the author's judgement, but I think it would be worth commenting on the distinction between "statistical multiplets" and multiplets arising from other sources.
+
+*Response:
+This is an excellent point, and similar to one made by Reviewer 1 about the validity of the Poisson assumption.
+I have added text to the Discussion (second paragraph) emphasizing how the calculation assumes a Poisson distribution which could be violated by factors such as cell clumping.
+However, I am not certain how to extend the calculations to identify deviations from Poisson behavior with the input data that the approach currently assumes.
+I think that an additional data point (such as the number of empty droplets) would need to be available in order to determine if the multiplet frequency is actually higher than expected under Poisson.
+We can say that non-Poisson behavior should make the calculations here conservative provided that there is not a bias towards clumping of cells of the same type; this fact is now noted in the Discussion.*

### 2018-08-04 paper.tex
completed revision

@@
 
-The most common method to determine the multiplet frequency is to use a mix two types of cells (e.g., human and mouse).
+The most common method to determine the multiplet frequency is to mix two types of cells (e.g., human and mouse).
 During the analysis of the sequencing results, each non-empty partition can be identified as containing transcripts from one or both of the two cell types.
@@
 However, in some cases the two cell types may be mixed in unequal proportions.
-Unequal mixing could arise simply from error during cell counting, but it can also be a desirable aspect of experimental design.
+Unequal mixing could arise simply from error during cell counting, or it could be an intentional aspect of experimental design~\citep{rosenberg2018single}.
 For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add fewer mouse cells so that most of the resulting data is for the human cells.
@@
 It is important to note that this calculation requires that the sequencing data have already been analyzed to determine whether each partition contains a non-negligible number of transcripts from each cell type, but many common analysis programs (such as the 10X \texttt{cellranger} pipeline) already do this.
-It is also important to note that the approach in this paper only calculates the multiplet frequency---it does \emph{not} actually identify the multiplets so that they can be removed from downstream analyses.
-For that purpose, other more sophisticated approaches have been developd~\citep{ilicic2016classification,stoeckius2017cell,kang2018multiplexed}.
+
+The calculation also assumes that the number of cells per droplets follows a Poisson distribution.
+While many single-cell RNA sequencing methods are designed to partition cells in a way that concords with this assumption~~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive} , it is possible that cell clumping or other factors could bias certain partitions to contain more cells than expected under a Poisson distribution.
+In such a scenario, the calculations in this paper would overestimate the true multiplet frequency if the clumping is equally likely across cell types, but could underestimate the true multiplet frequency if intra-cell-type clumping is more likely than inter-cell-type clumping.
+
+Finally, the approach in this paper only calculates the multiplet frequency---it does \emph{not} actually identify the multiplets so that they can be removed from downstream analyses.
+For that purpose, other more sophisticated approaches have been developd~\citep{ilicic2016classification,stoeckius2017cell,kang2018multiplexed,wolock2018scrublet,depasquale2018doubletdecon}.
 Nonetheless, simply calculating the multiplet frequency from the data returned by standard pipelines such as the 10X \texttt{cellranger} is important for many purposes, and the results here enable that to be done regardless of the proportions at which the cell types are mixed.

### 2019-02-26 paper.tex
fixed typo noticed by Lior Pachter

@@
 
-%\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
-\documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
+\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
+%\documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions
 
@@
 \Pr\left(c \ge 2\right) &=& 1 - \Pr\left(c = 0\right) - \Pr\left(c = 1\right) \nonumber \\
-&=& 1 - e^{-\mu_1 - \mu_2} - \left(\mu_1 + \mu_2\right) e^{-\mu_1 + \mu_2}.
+&=& 1 - e^{-\mu_1 - \mu_2} - \left(\mu_1 + \mu_2\right) e^{-\mu_1 - \mu_2}.
 \end{eqnarray}
@@
 M &=& \frac{\Pr\left(c \ge 2\right)}{\Pr\left(c \ge 1\right)} \nonumber \\
-&=& 1 - \frac{\left(\mu_1 + \mu_2\right) e^{-\mu_1 + \mu_2}}{1 - e^{-\mu_1 - \mu_2}}.
+&=& 1 - \frac{\left(\mu_1 + \mu_2\right) e^{-\mu_1 - \mu_2}}{1 - e^{-\mu_1 - \mu_2}}.
 \end{eqnarray}
@@
 The calculation also assumes that the number of cells per droplets follows a Poisson distribution.
-While many single-cell RNA sequencing methods are designed to partition cells in a way that concords with this assumption~~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive} , it is possible that cell clumping or other factors could bias certain partitions to contain more cells than expected under a Poisson distribution.
+While many single-cell RNA sequencing methods are designed to partition cells in a way that concords with this assumption~~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive}, it is possible that cell clumping or other factors could bias certain partitions to contain more cells than expected under a Poisson distribution.
 In such a scenario, the calculations in this paper would overestimate the true multiplet frequency if the clumping is equally likely across cell types, but could underestimate the true multiplet frequency if intra-cell-type clumping is more likely than inter-cell-type clumping.
